Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-05-09 second-draft/pnas-word-order-si.tex
revising

TODO

- about the regression

-- prior, how many samples, how many chains, convergence criteria

-- robustness to priors

@@

\section{Supporting Analyses for Study 1}

\paragraph{Other tests, maybe som viz}

\paragraph{Pareto Curve}

It has to be convex, so don't need to do more.

maybe per-language curves?

\section{Results on all UD Relations}
In this section, we provide the predicted prevalence of correlations between the \emph{obj} dependency and all UD dependency types, along with the expected prevalence according to typological studies.
We also report results for grammars optimized for predictability and parseability individually.

Further, we report results for grammars optimized for dependency length minimization (DLM).
DLM can be seen as a heuristic approximation to efficiency.
We also found that DLM is predicted by efficiency optimization.
Indeed, we show that optimizing for DLM makes predictions similar to efficiency optimization.

@@

@@

\section{Comparing Efficiency to its Components}

@@

\section{Possible values of $\lambda$}

@@

\subsection{Differentiable Ordering Grammars}

In order to carry out optimization of grammars, we extend the space of grammars to include probabilistic relaxations of grammars, which describe probability distributions over different ways of ordering a syntactic structure into a sentence.
This makes efficiency a differentiable function of the grammar parameters, and enables optimization with stochastic gradient descent, as we describe in Section~\ref{sec:optim-eff}.

@@

\textbf{rewrite this}

For each of our objective functions, we seek parameters $\theta$ of a word order grammar $L_\theta$ that maximize the average utility of sentences in the language obtained by linearizing unordered trees $t \sim T$ according to $L_\theta$. The calculation of the predictability and parseability objectives also requires fitting language models parameterized by $\phi$ and parsers parameterized by $\psi$, respectively; and the calculation of the efficiency objective requires joint optimization of the word order grammar, language models, and parsers. The optimization problems come out to:
\begin{align}
\max_\theta &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{DepL}(t; {\bf w}) \label{eq:maximize-depl}\\
\max_{\theta,\phi} &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{Pred}^{\phi(\theta)}(t; {\bf w}) \label{eq:maximize-pred}\\
\max_{\theta,\psi} &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{Pars}^{\psi(\theta)}(t; {\bf w}) \label{Eq:maximize-pars} \\
\max_{\theta,\phi,\psi} &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{Efficiency}^{\psi(\theta), \phi(\theta)}(t; {\bf w}). \label{eq:maximize-efficiency} 
\end{align}
Note that the original word orders from the treebank never enter this objective---in particular, they do not influence the language model and the parser, which are entirely determined by $\theta$.

The case of dependency length in Eq.~\ref{maximize-depl} is the simplest, requiring only coordinate descent on the parameters $a_\tau, b_\tau \in \theta$. A similar optimization problem was solved by hill-climbing in \citet{gildea2007optimizing} and \citet{gildea2015human}.
We optimize this objective by stochastic gradient descent with a batch size of one unordered dependency tree.

The case of predictability, parseability, and efficiency are more challenging, as we need to simultaneously look for an ordering model optimized for the loss of a language model and/or a parser, and for a language model and/or parser that are optimized for that given ordering model.
We solve Eqs.~\ref{eq:maximize-pred}--\ref{eq:maximize-efficiency} by performing stochastic gradient descent over all sets of parameters $\theta, \phi$, and/or $\psi$ simultaneously.
Below we discuss the process only for the predictability objective (Eq.~\ref{eq:maximize-pred}); the other objectives are optimized analogously.
In each step, we sample a dependency tree $t$ from the treebank, then sample an ordering from the current setting of $\theta$ to obtain a linearized sentence ${\bf w} \sim P_{L_\theta}(\cdot|t)$.
Then we %update $\thetal$ with ordinary gradient descent, and $\thetad$ with the REINFORCE estimator~\ref{williams-simple-1992}.
do a gradient descent step using the estimator %(analogous for $R_{Pars}$ instead of $R_{Pred}$)
\begin{equation}\label{eq:estimator}
{ \partial_\theta \left(\log P_{L_\theta}({\bf w}|t)\right) \cdot R^\phi_{Pred} (t; {\bf w}) \choose  \partial_\phi R_{Pred}^\phi(t; {\bf w})}
%{- \partial_\thetad \left(\log P_\thetad({\bf w})\right) \cdot R_{Pred}(\bf{w}, \thetal) \choose  - \partial_\thetal \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i})}
\end{equation}
%$${- \partial_\thetad \left(\log P_\thetad({\bf w})\right) \cdot \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i}) \choose  - \partial_\thetal \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i})}$$
for the gradient
$${ \partial_\theta \choose \partial_\phi} \E_{{\bf w'} \sim L_\theta(t)} R_{Pred}^\phi(t; {\bf w'}).$$
This unbiased estimator is the ordinary gradient estimator for $\phi$, and the REINFORCE estimator for $\theta$ \citep{williams1992simple}.

In addition, we estimated maximum-likelihood ordering grammars on the original ordered dependency trees with SGD. For nonprojective trees, we ignore discontinuities.
%and $\ell_2$ regularization.

A different commonly considered information-theoretic objective function for word order is Uniform Information Density~\citep{jaeger2010redundancy}, formalized as minimizing a superlinear function of per-word surprisal~\citep{levy:2018cogsci}.
Note that optimizing grammars for this objective faces additional challenges, as the optimization problem cannot be rewritten as joint optimization in the form (\ref{eq:maximize-pred}).
We thus do not evaluate this objective function here.

@@

\textbf{TODO study 1 only has 1 set of parameters}

@@

@@

@@

Here we address this question by demonstrating that our languages optimized for prediction and parseability are also optimal under a range of different language models and parsers. These results show that our optimization process creates language in which strings are generally predictable and informative about trees, without dependence on a particular prediction and parsing algorithms.

@@

Many of the word order biases explained by parseability in Table~\ref{tab:result-dryer} are also explained by dependency length minimization. Therefore, we address here the idea that the graph-based parser might have a built-in bias toward parses involving short dependencies, which we call a \key{locality bias}.

A simple way to eliminate or alter a locality bias for any parser would be to change the order in which the parser sees words. Suppose that a parser $P$ has a bias toward positing a dependency between a word and the immediate previous word.

@@

@@

We constructed a Bigram model with Kneser-Ney smoothing.
A bigram model predicts each word taking only the previous word into account.
This contrasts with LSTMs, which take the entire context into consideration.
Thus, bigram models and LSTMs stand on opposing ends of a spectrum of language models taking more and more aspects of the context into account.

@@

\subsection{Discussion}

It is always a logical possibility in studies such as this one that the results are dependent on the particular probabilistic models used.

@@

\textbf{do same for language model, and for the other random grammars}

@@

\begin{figure}
    \centering
    \includegraphics[scale=.4]{figures/learning-parser-czech.pdf} 
        \caption{Parsing accuracy (measured in UAS) for optimized (light blue) and random (black) ordering grammar on Czech data, as a function of the fraction of total training data provided.}
    \label{fig:learning-czech}
\end{figure}

>>>>

\tableofcontents

\section{Study 2: Regression for Predicted Correlations}

Following (CITE), we used as a very weakly informative prior a Student's $t$ prior with 3 degrees of freedom, mean 0, and scale $\sigma=10$ (i.e., the PDF $p$ is $\frac{1}{\sigma} p_3(x/\sigma)$).
The covariance matrix is modeled using the LKJ prior, as described in (CITE brms Bayesian multilevel).
We used four chains, with 5000 samples each (2500 warmup).
The chains had converged ($\hat{R} < 1.01$).

To verify robustness to more informative priors, we conducted the analysis (REF) with a tighter weakly informative prior.
Note that an intercept $\alpha \approx 2.19$ in the logistic model corresponds to a correlation that holds in 90\% of cases; we thus selected the prior $\mathcal{N}(0,2)$.
We obtained a qualitatively equivalent result (posterior probability $0.932$ of predicting all eight correlations).
We obtained qualitatively equivalent results for analyses on the levels of individual relations when estimating the model using \texttt{lme4} (CITE) and obtaining $p$-values from the resulting fit: For all eight correlations, the sign of $\alpha$ was positive ($p < 0.05$).

@@

\section{Study 1: Supporting Analyses for Study 1}

- binomial test

\paragraph{z-Scored Data (Figure (REF))}
We z-transformed on the level of individual languages, normalizing the mean and SD parseability and predictability of the (1) real grammar, (2) the mean of predictability and parseability of all random grammars, (3) the grammar optimized for efficiency (at $\lambda =0.9$, see Section REF), (4) grammar optimized for parseability only, and (5) grammar optimized for predictability only.

Within each language, we estimate the Pareto curve by connecting the points obtained by optimizing for (1) efficiency (at $\lambda = 0.9$, see Section REF), (2) parseability, and (3) predictability.
We then estimated the overall average Pareto curve, as shown in Figure (REF), by averaging the z-transformed points (1), (2), (3).

\section{Study 2: Results on all UD Relations}
In this section, we provide the predicted prevalence of correlations between the \emph{obj} dependency and all UD dependency types, along with the expected prevalence according to typological studies.
We also report results for grammars optimized for predictability and parseability individually.

@@

Further, we report results for grammars optimized for dependency length minimization (DLM).
Following (CITE), we formalize the dependency length of a sentence as the sum of the distances of all pairs of syntactically related words, where distance is measured by the number of intervening words.
Prior work has suggested principles akin to DLM as approximating efficiency optimization of grammars (CITE).
Indeed, we found that DLM is predicted by efficiency optimization, as reported in Discussion (Main Paper).
In line with these ideas, we show that optimizing for DLM makes predictions similar to efficiency optimization, itself a novel result.

@@

\section{Study 2: Comparing Efficiency to its Components}\label{sec:posterior-number}

In Figure~\ref{fig:posterior}, we plot the posterior number of correlations predicted to hold in most optimized grammars, as obtained from the Bayesian regression.
In addition to grammars optimized for efficiency, we also report the result for grammars optimized for predictability and for parseability alone.
Efficiency predicts all eight correlations with high posterior probability; predictability and parseability alone do not.

@@

\section{Possible values of $\lambda$}\label{sec:lambda}

@@

In this section, we describe the method we employ for creating grammars that are optimized for efficiency.
Our main methodological innovation is to carry out optimization in an extended space of grammars that interpolates continuously between different grammars (Section~\ref{sec:diff-gramm}).
More specifically, we include probabilistic relaxations of grammars, which describe probability distributions over different ways of ordering a syntactic structure into a sentence.
This makes efficiency a \emph{differentiable} function of the grammar parameters, and enables efficient optimization with stochastic gradient descent, as we describe in Section~\ref{sec:optim-eff}.

This method addresses a major challenge noted in previous work optimizing grammars, namely that the predictability (and parseability) of an individual sentence depends on the entire distribution of the language.
Previously, \citep{gildea2015human} optimized grammars for dependency length and trigram surprisal using a simple hill-climbing method on the grammar parameters, which required reestimating the trigram surprisal model in every iteration.
Such a method would be computationally prohibitive for efficiency optimization, as it would require reestimating the neural network models after every change to the grammar, which would amount to reestimating them hundreds or thousands of times per grammar.
Our method, by allowing for the use of stochastic gradient descent, addresses this challenge, as we describe in Section~\ref{sec:optim-eff}.

\subsection{Differentiable Ordering Grammars}\label{sec:diff-gramm}

@@

A word order grammar can be viewed as a function $\mathcal{L}_\theta$, whose behavior is specified by parameters $\theta$, which takes an unordered dependency tree $t$ as input and produces as output an ordered sequence of words $\mathcal{L}_\theta(t) = \mathbf{w}$ linearizing the tree.
More generally, if $\mathcal{L}_\theta$ is a differentiable ordering grammar (Section~\ref{sec:diff-gramm}), then $\mathcal{L}_\theta(t)$ is a \emph{probability distribution} over ordered sequences of words $\mathbf{w}$.
In the limit where $\mathcal{L}_\theta$ becomes deterministic, the distribution $\mathcal{L}_\theta(t)$ concentrates on a single ordering.

Let $T$ be a distribution over unordered dependency trees and let $L_\theta$ be a differentiable word order grammar. Our goal is to find parameters $\theta$ which maximize an objective function $J$ in expectation over utterances:
\begin{equation}
\label{eq:exp-over-utts}
J_T(\theta) = \E_{t \sim T} \E_{\mathbf{w} \sim \mathcal{L}_\theta(t)} [R(t; \mathbf{w})], % TODO make the expectations look better!!
\end{equation}
where the function $R$ specifies a score for an individual utterance $\mathbf{w}$ expressing a tree $t$. In practice, $T$ will be the empirical distribution over unordered trees observed in a dependency treebank.
Below we discuss possible values of $R$, which may implement predictability, parseability, or a weighted combination of the two, representing overall efficiency.

We formalize predictability as follows. Given a language model $P_\phi$ with parameters $\phi$, the log probability of a sentence ${\bf w}$ is:
\begin{equation*}
R_{Pred}^\phi(t; {\bf w}) = \sum_{i=1}^{\#{\bf w}} \log P_\phi(w_{i+1}|w_{1\dots i}),
\end{equation*}
where the language model parameters $\phi$ are chosen predict the language induced by tree distribution $T$ and word order grammar $\mathcal{L}_\theta$ as accurately as possible (maximizing predictability):
\begin{equation*}
	\phi(\theta) = \argmax_\phi \displaystyle \E_{t \sim T} \E_{{\bf w} \sim L_\theta(t)} [ R^\phi_{Pred}({\bf w}, \phi) ].
\end{equation*}

We specify parseability as the mutual information between strings and trees, or equivalently the quantity of information provided by strings about trees. Formally, we say the parseability score of an utterance $\mathbf{w}$ is the pointwise mutual information of the $\mathbf{w}$ and the tree $t$ it was generated from, calculated using a parser with parameters $\psi$:\footnote{Note that $P(t)$ is a constant term which will be ignored in the optimization process.}
\begin{align}
\nonumber
R_{Pars}^\psi(t; {\bf w}) &= \log \frac{P_\psi (t|\mathbf{w})}{P(t)} \\
\nonumber
&= \displaystyle \sum_{i=1}^{\#\mathbf{w}} \log P_\psi\left(\begin{array}{@{}l@{}}head(w_{i}) \\ label(w_i)\end{array}{\big |}\textbf{w}\right) - \log P(t),
\end{align}
where the parser parameters $\psi$ are chosen to maximize parsing accuracy:
\begin{equation*}
\psi(\theta) = \argmax_\phi \displaystyle \E_{t \sim T} \E_{{\bf w} \sim L_\theta(t)} [ R_{Pars}({\bf w}, \psi) ].
\end{equation*}

For efficiency, we linearly combine the two terms:
\begin{equation}
\label{eq:combined-score}
R_{Efficiency}^{\phi,\psi}(t; {\bf w}) = R_{Pars}^\psi(t; {\bf w}) + \lambda R_{Pred}^\phi(t; {\bf w}),
\end{equation}

%\begin{equation}
%    \label{eq:depl}
%	R_{DepL}(t; {\bf w}) = -\sum_{i=1}^{\# {\bf w}} |i - head(t; w_i)|,
%\end{equation}
%where $head(t; i) \in \{1, ..., n\}$ is the index of the head of $w_i$, and the sum is negative because dependency length is seen as a cost.

For each of our objective functions, we seek parameters $\theta$ of a word order grammar $\mathcal{L}_\theta$ that maximize the average utility of sentences in the language obtained by linearizing unordered trees $t \sim T$ according to $\mathcal{L}_\theta$. The calculation of the predictability and parseability objectives also requires fitting language models parameterized by $\phi$ and parsers parameterized by $\psi$, respectively; and the calculation of the efficiency objective requires joint optimization of the word order grammar, language models, and parsers. The optimization problems come out to:
\begin{align}
%\max_\theta &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{DepL}(t; {\bf w}) \label{eq:maximize-depl}\\
\max_{\theta,\phi} &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{Pred}^{\phi(\theta)}(t; {\bf w}) \label{eq:maximize-pred}\\
\max_{\theta,\psi} &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{Pars}^{\psi(\theta)}(t; {\bf w}) \label{Eq:maximize-pars} \\
\max_{\theta,\phi,\psi} &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{Efficiency}^{\psi(\theta), \phi(\theta)}(t; {\bf w}). \label{eq:maximize-efficiency} 
\end{align}
Note that the original word orders from the treebank never enter this objective---in particular, they do not influence the language model and the parser, which are entirely determined by $\theta$.

%The case of predictability, parseability, and efficiency are more challenging, as 
%e need to simultaneously look for an ordering model optimized for the loss of a language model and/or a parser, and for a language model and/or parser that are optimized for that given ordering model.
We solve Eqs.~\ref{eq:maximize-pred}--\ref{eq:maximize-efficiency} by performing stochastic gradient descent over all sets of parameters $\theta, \phi$, and/or $\psi$ simultaneously.
Below we discuss the process only for the predictability objective (Eq.~\ref{eq:maximize-pred}); the other objectives are optimized analogously.

We initialize all values $a_\tau := 0.5$, $b_\tau := 0$ (except for the \emph{obj} dependency, for which we fix $a_\tau$ to 0 or 1, see Section (REF)).

In each optimization step, we sample a dependency tree $t$ from the treebank, then sample an ordering from the current setting of $\theta$ to obtain a linearized sentence ${\bf w} \sim P_{L_\theta}(\cdot|t)$.
Then we %update $\thetal$ with ordinary gradient descent, and $\thetad$ with the REINFORCE estimator~\ref{williams-simple-1992}.
do a gradient descent step using the estimator %(analogous for $R_{Pars}$ instead of $R_{Pred}$)
\begin{equation}\label{eq:estimator}
{ \partial_\theta \left(\log P_{L_\theta}({\bf w}|t)\right) \cdot R^\phi_{Pred} (t; {\bf w}) \choose  \partial_\phi R_{Pred}^\phi(t; {\bf w})}
%{- \partial_\thetad \left(\log P_\thetad({\bf w})\right) \cdot R_{Pred}(\bf{w}, \thetal) \choose  - \partial_\thetal \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i})}
\end{equation}
%$${- \partial_\thetad \left(\log P_\thetad({\bf w})\right) \cdot \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i}) \choose  - \partial_\thetal \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i})}$$
for the gradient
$${ \partial_\theta \choose \partial_\phi} \E_{{\bf w'} \sim L_\theta(t)} R_{Pred}^\phi(t; {\bf w'}).$$
This unbiased estimator is the ordinary gradient estimator for $\phi$, and the REINFORCE estimator for $\theta$ \citep{williams1992simple}.

See Section (REF) for the stopping criterion.

%The case of dependency length in Eq.~\ref{maximize-depl} is the simplest, requiring only coordinate descent on the parameters $a_\tau, b_\tau \in \theta$. A similar optimization problem was solved by hill-climbing in \citet{gildea2007optimizing} and \citet{gildea2015human}.
%We optimize this objective by stochastic gradient descent with a batch size of one unordered dependency tree.

%In addition, we estimated maximum-likelihood ordering grammars on the original ordered dependency trees with SGD. For nonprojective trees, we ignore discontinuities.
%and $\ell_2$ regularization.

%A different commonly considered information-theoretic objective function for word order is Uniform Information Density~\citep{jaeger2010redundancy}, formalized as minimizing a superlinear function of per-word surprisal~\citep{levy:2018cogsci}.
%Note that optimizing grammars for this objective faces additional challenges, as the optimization problem cannot be rewritten as joint optimization in the form (\ref{eq:maximize-pred}).
%We thus do not evaluate this objective function here.

@@

%\textbf{TODO study 1 only has 1 set of parameters}

@@

\paragraph{Optimized Grammars}

@@

For efficiency optimization, for each grammar, we ran efficiency optimization with three different random seeds, selecting among these the seed that yielded the best overall efficiency value.
This

@@

Here we address this question by demonstrating that our languages optimized for efficiency are also optimal under a range of different language models and parsers. These results show that our optimization process creates language in which strings are generally predictable and informative about trees, without dependence on particular prediction and parsing algorithms.

@@

In this section, we address the idea that the graph-based parser might have a built-in bias toward certain kinds of orderings, and the question whether this might be responsible for our findings.
In particular, we address the idea that the graph-based parser might have a bias toward parses involving short dependencies, which we call a \key{locality bias}. 
We address this by changing the order in which the parser sees words, in such a way that the distance between words in the input is not indicative of syntactic relations.

%In particular, this allows us to show tha

%Many of the word order biases explained by parseability in Table~\ref{tab:result-dryer} are also explained by dependency length minimization. Therefore, we address here the idea that the graph-based parser might have a built-in bias toward parses involving short dependencies, which we call a \key{locality bias}.
%A simple way to eliminate or alter a locality bias for any parser would be to change the order in which the parser sees words. Suppose that a parser $P$ has a bias toward positing a dependency between a word and the immediate previous word.

@@

The result is that distance between words in the input is not indicative at all of the presence of absence of syntactic relations between them.

@@

We model predictability using LSTM language models, which are are the strongest known predictors of the surprisal effect on human processing effort~\citep{frank2011insensitivity,goodkind2018predictive}.
In previous work, such as \cite{gildea2015human}, predictability has often been measured using n-gram models.

Here, we show that languages optimized for LSTM predictability are also optimal for n-gram predictability.
Specifically, we constructed bigram models with Kneser-Ney smoothing (CITE).
A bigram model predicts each word taking only the previous word into account.
This contrasts with LSTMs, which take the entire context into consideration.
Thus, bigram models and LSTMs stand on opposing ends of a spectrum of language models taking more and more aspects of the context into account.

@@

%\subsection{Discussion}
%
%It is always a logical possibility in studies such as this one that the results are dependent on the particular probabilistic models used.

@@

\textcolor{red}{do same for language model, and for the other random grammars}

@@

\begin{figure}
    \centering
    \includegraphics[scale=.4]{figures/learning-parser-czech.pdf} 
        \caption{Parsing accuracy (measured in UAS) for optimized (light blue) and random (black) ordering grammar on Czech data, as a function of the fraction of total training data provided.\textcolor{red}{font size}}
    \label{fig:learning-czech}
\end{figure}
